Guidelines for reliable and reproducible functional enrichment analysis

### 2021-09-02 manuscript.Rmd
editing

Source: https://github.com/markziemann/SurveyEnrichmentMethods

Target Journal: Briefings in Bioinformatics (JIF=9.0)

@@

The purpose of this work is to survey the frequency of methodological and reporting flaws in the literature, in particular;
(i) lack of a background gene set, 
(ii) lack of p-value adjustment for multiple comparisons and 
(ii) lack of essential methodological details.
We use the results of this survey to inform a set of minimum standards for functional enrichment analysis.

@@

We collated 2941 articles in PubMed Central published in 2019 that have keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
We sampled 200 of these articles and collected the following information from the article, searching the methods sections and other parts of the article including the supplement.

@@

| 1 point deducted | 1 point awarded |
| --- | --- |
| Gene set library origin not stated | Code made available |
| Gene set library version not stated | Gene lists provided |
| Stat test not stated | |
| No stat test conducted | |
| No FDR correction conducted | |
| App used not stated | |
| App version not stated | |
| Background list not defined | |
| Inappropriate background list used | |

Scimago Journal Rank (SJR) data were downloaded from the Scimago website (https://www.scimagojr.com/journalrank.php) and used to score journals by their citation metrics.
Using NCBI’s Eutils API, we collected the number of citations each article accrued since publication.
We used Pearson correlation tests to assess the association with the analysis scores we generated.
All data analyses were conducted in R v4.1 and the analysis scripts are available at GitHub (https://github.com/markziemann/SurveyEnrichmentMethods).

@@

In order to demonstrate the effect of functional enrichment analysis misuse, we used an example RNA-seq dataset examining the effect of high glucose exposure on hepatocytes.
Out of 39,297 genes in the annotation set, 15,635 were above the detection threshold (>10 reads per sample on average).
Statistical analysis revealed 3,472 differentially expressed genes with 1,560 up-regulated and 1,912 down-regulated (FDR<0.05) due to glucose exposure.

@@

![Figure 1. Example of enrichment analysis misuse. (A) Comparison of FCS and ORA methods. (B) ORA with standard and whole genome background. (C) Effect of combining up and downregulated genes on ORA results. ](images/Figure1.png "Figure 1")

@@

A search of PubMed Central showed 2941 articles published in 2019 with the keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
From these, 200 articles were screened for methodological errors.
We excluded 14 articles from the screen because they did not present any enrichment analysis. 
Those excluded articles included articles describing novel enrichment analysis techniques or tools, review articles or conference abstracts.
As some articles included more than one enrichment analysis, the dataset included 235 analyses from 186 articles; this data is available in Table S1.
A flow diagram of the survey is provided in Figure 2.

@@

For analyses using ORA methods, we studied what background gene set was used (Figure 3G). 
This revealed that in most cases, the background list was not stated or it was clear from the article methods section that no background list was used.
In a few cases, background list was mentioned but was inappropriate, for example using a whole genome background for an assay like RNA-seq.
In only 8/197 of cases (4.1%), the appropriate background list was described in the article.

@@

In order to make an accurate assessment of association between analysis quality and bibliometrics, we required a larger sample of articles.
We therefore analysed a further 1300 articles, bringing the total number of analyses described up to 1630; this dataset is available in Table S2.

@@

The proportion of ORA studies that combined up- and down-regulated gene set was also surprisingly high, at 70% (Figure 3K).
As exemplified in Figure 1C, combining up- and down-regulated genes can severely reduce the number of significant  gene sets (from 293 pathways to just 22 in this example).
This is consistent with a previous report indicating that combining up- and down-regulated gene lists is considered poor practice because it reduces sensitivity (Hong et al, 2014).
These results suggest that a large fraction of published works suffer from a high rate of false negatives, in effect missing a large fraction of truly differentially regulated gene sets.
This is exacerbated by using an inappropriate background gene list, which leads to a high rate of false positives.

Although analyses involving GSEA scored better overall, they were not free of issues.
For example GSEA has different options that impact the results including the ranking metric, gene weighting method and the permutation type (on samples or genes), which were not regularly reported in articles (Figure 3L).

@@

Still there are some limitations that need to be recognised.
Many open access articles examined here are from journals not considered to be highly prestigious, and the distribution of analysis scores might be different for closed access articles.
The articles included in this study contained keywords related to functional enrichment in the abstract, and it is plausible that articles in higher ranked journals contain such details in the abstract at lower rates.

@@

3. If making claims about the regulation of a gene set, then results of a statistical test must be shown (including measures of significance and effect size).
The number of genes belonging to a pathway on its own is insufficient to infer enrichment.
"Data not shown" is not acceptable when it is possible to attach supplementary files.

@@

7. Report selection criteria and parameters.
If performing ORA, then the inclusion thresholds need to be mentioned.
If using GSEA or another FCS tool, parameters around ranking, gene weighting and type of test need to be disclosed (eg: permuting sample labels or gene labels).
Any parameters that are vary from the defaults should be reported.

@@

### Additional recommendations for enhanced reproducibility

>>>>

@@

The purpose of this work is to survey the frequency of methodological and reporting flaws in the literature, in particular;
(i) lack of a background gene set, 
(ii) lack of p-value adjustment for multiple comparisons and 
(iii) lack of essential methodological details.
We use the results of this survey to inform a set of minimum standards for functional enrichment analysis.

@@

We collated 2,941 articles in PubMed Central published in 2019 that have keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
We sampled 200 of these articles and collected the following information from the article, searching the methods sections and other parts of the article including the supplement.

@@

*Table 1. Scoring schema.*
| 1 point deducted | 1 point awarded |
| --- | --- |
| Gene set library origin not stated | Code made available |
| Gene set library version not stated | Gene profile data provided |
| Statistical test not stated | |
| No statistical test conducted | |
| No FDR correction conducted | |
| App used not stated | |
| App version not stated | |
| Background list not defined | |
| Inappropriate background list used | |

Scimago Journal Rank (SJR) data for 2020 were downloaded from the Scimago website (https://www.scimagojr.com/journalrank.php) and used to score journals by their citation metrics.
Using NCBI’s Eutils API, we collected the number of citations each article accrued since publication.
We used Pearson correlation tests to assess the association with the analysis scores we generated.
All data analyses were conducted in R v4.1 and the analysis scripts are available at GitHub (https://github.com/markziemann/SurveyEnrichmentMethods).

@@

In order to demonstrate the effect of functional enrichment analysis misuse, we used an example RNA-seq dataset examining the effect of high glucose exposure on hepatocytes.
Out of 39,297 genes in the annotation set, 15,635 were above the detection threshold (>10 reads per sample on average).
Statistical analysis revealed 3,472 differentially expressed genes with 1,560 up-regulated and 1,912 down-regulated (FDR<0.05) due to elevated glucose exposure.

@@

![Figure 1. Example of enrichment analysis misuse. Euler diagrams show the overlap between sets of differentially regulated gene sets with different methods (FDR<0.05). (A) Comparison of FCS and ORA methods. (B) ORA with standard and whole genome background. Whole genome background analysis is indicated with a \* (C) Effect of combining up and downregulated genes on ORA results. Analysis using combined gene list is indicated by *ORA comb*, while *ORA\* comb* indicates combined gene list analysis with a whole genome background. ](images/Figure1.png "Figure 1")

@@

A search of PubMed Central showed 2,941 articles published in 2019 with the keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
From these, 200 articles were selected for methodological analysis.
We excluded 14 articles from the screen because they did not present any enrichment analysis. 
Those excluded articles included articles describing novel enrichment analysis techniques or tools, review articles or conference abstracts.
As some articles included more than one enrichment analysis, the dataset included 235 analyses from 186 articles; this data is available in Table S1.
A flow diagram of the survey is provided in Figure 2.

@@

For analyses using ORA methods, we studied what background gene set was used (Figure 3G). 
This revealed that in most cases, the background list was not defined or it was clear from the article methods section that no background list was used.
In a few cases, a background list was mentioned but was inappropriate, for example using a whole genome background for an assay like RNA-seq.
In only 8/197 of cases (4.1%), the appropriate background list was described in the article.

@@

In order to make an accurate assessment of association between analysis quality and bibliometrics, we required a larger sample of articles.
We therefore analysed a further 1,300 articles, bringing the total number of analyses described up to 1630; this dataset is available in Table S2.

@@

The proportion of ORA studies that combined up- and down-regulated gene list prior to analysis was also surprisingly high, at 70% (Figure 3K).
As exemplified in Figure 1C, combining up- and down-regulated genes can severely reduce the number of significant  gene sets (from 293 pathways to just 22 in this example).
This is consistent with a previous report indicating that combining up- and down-regulated gene lists is considered poor practice because it reduces sensitivity (Hong et al, 2014).
These results suggest that a large fraction of published works suffer from a high rate of false negatives, in effect missing a large fraction of truly differentially regulated gene sets.
This is exacerbated by using an inappropriate background gene list, which leads to a high rate of false positives.

Although analyses involving GSEA scored better overall, they were not free of issues.
For example GSEA has different options that impact the results including the ranking metric, gene weighting method and the permutation type (on samples or genes), which were not regularly reported in articles (Figure 3L), limiting reproducibility.

@@

Still there are some limitations that need to be recognised.
Many open access articles examined here are from lower-ranked journals, and might not be representative of articles in paywalled journals.
The articles included in this study contained keywords related to functional enrichment in the abstract, and it is plausible that articles in higher ranked journals contain such details in the abstract at lower rates.

@@

3. If making claims about the regulation of a gene set, then results of a statistical test must be shown including measures of significance and effect size.
The number of genes belonging to a pathway on its own is insufficient to infer enrichment.

@@

7. Report selection criteria and parameters.
If performing ORA, then the inclusion thresholds need to be mentioned.
If using GSEA or another FCS tool, parameters around ranking, gene weighting and type of test need to be disclosed (eg: permuting sample labels or gene labels).
Any parameters that vary from the defaults should be reported.

@@

### Additional recommendations for enhanced reliability and reproducibility
